Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins

### 2017-06-07 Manuscript.tex
Minor final changes

\begin{abstract}
We consider a generalized-ensemble algorithm for coarse-grained simulations of biomolecules which allows the thermodynamic behavior of two or more sequences to be determined in a single ``multisequence" run. By carrying out a random walk in sequence space, the method also enhances conformational sampling. Escape from local energy minima is accelerated by visiting sequences for which the minima are more shallow or absent. We apply the method alongside an intermediate-resolution coarse-grained model for protein folding with 7 atoms per amino acid and 3 amino acid types. The potential for large-scale coverage of sequence space is explored by applying the method to two sets with $>$1,000 sequences each. The resulting thermodynamic data is used to explore the sequence-structure and sequence-stability relationships between pairs of protein folds with different secondary structures.
\end{abstract}

@@

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, many biological processes naturally call for the investigation and comparison of molecular variants. This is required, for example, in determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} or protein-nucleotide interactions,~\cite{Rohs2010} and the role of mutations in disease processes such as protein aggregation.~\cite{Ross2004} Another example is the protein folding process, where unique insight has been achieved by comparing the folding within and between protein families.~\cite{Tzul2017,Wensley2010} In a situation with rapid growth of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

To this end, we consider in this work an MC-based algorithm that can calculate the thermodynamics of multiple sequences in a single run and apply it to a coarse-grained model for protein folding. This multisequence (MS) method was originally developed in the context of homo- and heteropolymer simulations~\cite{Irback1995} and was also later adapted for the characterization of peptide-protein binding specificity.~\cite{Bhattacherjee2013,Wallin2017} To our knowledge, however, it has not been previously tested in realistic protein folding simulations. The MS algorithm carries out a simulation in a generalized ensemble that performs a random walk in sequence space. Hence, there are two main types of updates: conformational updates $\rbar\rightarrow\rbar'$ and sequence updates $\sbar\rightarrow\sbar'$. As illustrated in Fig.~1, this strategy can be applied when $\rbar$ and $\sbar$ are ``perpendicular" coordinates such that the potential energy of the model can be written in terms of two independent variables, $E(\sbar,\rbar)$. While this may be hard to achieve for detailed biomolecular models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model protein sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and a $\beta$-hairpin, respectively.~\cite{Bhattacherjee2012} In a previous work, this set allowed us to demonstrate the existence of A1-N1 mutational pathways which do not proceed through disordered (non-folding) sequences but instead switch abruptly between the two folds in a single mutational step.~\cite{Holzgrafe2014} Moreover, it allowed the exploration of the biophysical factors that underpin this ``fold switching" phenomenon,~\cite{Holzgrafe2015} which was recently demonstrated in a handful of natural and engineered proteins.~\cite{Bryan2010} Here we use the set {\SR} and our CG model for protein folding to validate the MS method and compare its efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences, also spanning the A1-N1 space, as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively. Besides showing that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical character of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

@@

\subsection{Generalized-ensemble algorithms and simulated tempering}
\noindent
Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters, such that a more efficient random walk in potential energy can be achieved.

A well-known generalized-ensemble algorithm is simulated tempering (ST),~\cite{Marinari1992,Lyubartsev1992} in which it is the temperature $T$ that becomes a dynamic parameter. In this scheme, frequent visits to high-$T$ allow simulations to readily escape from local energy traps. The ST algorithm thus simulates the joint probability distribution 
\begin{equation}
P(\rbar,m) =\dfrac{1}{\hat{Z}} e^{-\beta_m E(\rbar) + g_m}\,,
\label{ST}
\end{equation}
where  $\beta_m=1/k_\mathrm{B} T_m$, $\{T_m\}_{m=1}^\mathrm{M}$ a set of temperatures and $k_\mathrm{B}$ is Boltzmann's constant. The normalization constant in Eq.~\ref{ST} is  
\begin{equation}
\hat{Z} = \sum_r \sum_{m=1}^{\mathrm{M}}e^{-\beta_m E(\rbar) + g_m}\,,
\end{equation}
where the first sum is over all conformations $\rbar$. The simulation parameters $g_m$ control the marginal probability distribution
\begin{equation}
P(m) = \frac{1}{\hat{Z}}\sum_r e^{-\beta_m E(\rbar) + g_m} \,,
\end{equation}
and must therefore be carefully chosen. A common and convenient choice is $g_m\approx \beta_m F_m$, where $F_m$ is the free energy at temperature $T_m$. With this choice, $P(m)$ becomes approximately flat ensuring all temperatures are frequently visited.

\subsection{Multisequence algorithm}
\noindent 
The basic idea of the MS algorithm for biomolecular simulation is to let the sequence $\sbar$ become a dynamic parameter rather than the temperature as in ST. A dynamic $\sbar$ is technically feasible as long as the potential energy can be written as $E(\sbar,\rbar)$, where $\sbar$ and $\rbar$ are independent variables. This is the case in our coarse-grained protein model with only backbone degrees of freedom and can also be achieved for more detailed  models~\cite{Bhattacherjee2013,Wallin2017}.

In analogy with ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{\tilde{Z}}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
\tilde{Z} = \sum_{\sbar}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\end{equation}
and the first sum goes over a set of allowed sequences $\sbar$. The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=\tilde{Z}^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

@@

This geometrical description is paired with a simplified but finely tuned energy function with 4 terms: $E= E_\mathrm{ev} + E_\mathrm{loc} + E_\mathrm{hb} + E_\mathrm{hp}$. The first two, $E_\mathrm{ev}$ and $E_\mathrm{loc}$, represent excluded-volume effects of all atoms and local electrostatic effects, respectively. The hydrogen-bond energy, $E_\mathrm{hb}$, represents directionally dependent interactions between NH and CO groups which are necessary for secondary structure formation. Finally, the ``hydrophobicity" term, $E_\mathrm{hp}$, implements pairwise Lennard-Jones-like interactions between the $C_\beta$ atoms of h amino acids which are necessary for driving chain collapse during folding. Various model parameters, e.g., the strengths of hydrophobic attractions and hydrogen bonding, were determined based on the ability of the model to spontaneously fold a set of model sequences with 18-54 amino acids into  structurally diverse and thermodynamically stable native states with both $\beta$ and $\alpha$-structure. As it turned out, this strategy made the model robust enough to fold sequences designed to have mixed $\alpha$ and $\beta$ structures.

@@

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and described briefly below. First we define two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating the fraction of the fold-specific backbone-backbone hydrogen bonds that have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The  stabilities of folds IA and IB are then defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.80)$, respectively, i.e., the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ thus depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.875\pm 0.003$ for A1 and $\PIB=0.785\pm0.008$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $\mathrm{C}_\beta$-$\mathrm{C}_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} In analogy with $\PIA$ and $\PIB$, we define the stabilities of folds IIA and IIB as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.80)$, respectively. The root-mean-square-deviation, RMSD, is calculated over all $\mathrm{C}_\alpha$ atoms.

\begin{figure}
\includegraphics[width=10.0cm]{MCTrajFolds}
\caption{(A) Example of an MS simulation applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$. (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

@@

\subsection{Computational efficiency}
\noindent
We start by applying the MS algorithm to the set {\SR} across a range of temperatures $T$ (see Table~II). Two of the sequences in {\SR} are A1 and N1 (see Table~I) which fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. A1 and N1 differ at 10 positions such that 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences in {\SR} were selected from this binary space with the constraints that the total number of hydrophobic amino acids are not too high and that they are not too unevenly distributed.~\cite{Holzgrafe2014}

\begin{figure}
\includegraphics[width=7.8cm]{Pacc}
\caption{Acceptance rates for $\sbar\rightarrow\sbar'$ updates in MS simulations of the {\SR} sequence set as a function of (A) the number of changed amino acid positions $\Delta h$ and (B) temperature $T$. Acceptance rates for 3 different $T$'s are shown in (A).}
\end{figure}

Figure~2 illustrates a typical MS simulation trajectory carried out at the lowest studied temperature which is below the folding temperature of both A1 and N1.~\cite{Holzgrafe2014,Holzgrafe2015} From the MC evolution of the total energy $E$, sequence index $\sbar$, and RMSD values from the representative structures in Fig.~2B, it is evident that visits to various sequences drive transitions into a range of structural states. In particular, there are frequent visits to both $\alpha$-helix and $\beta$-hairpin structures and transitions between them are accompanied by a shift in which sequences are preferably visited. For example, visits to high $\sbar$-indices, including N1 with index 144, tend to coincide with formation of $\beta$-hairpin structures as required to generate the correct equilibrium conformational ensembles.

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rate is substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~4B). We note therefore that increased acceptance rates can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains.

\begin{figure}
\includegraphics[width=8.0cm]{Stderr}
\caption{Comparing sampling efficiency of the MS and ST algorithms. Statistical errors $\sigma_{\left < E\right >}$ of the average total energy $\left < E\right >$ from the two methods for the 16-amino acid sequences (A) A1, (B) N1, (C) R1 and (D) R2. Simulation lengths in the two methods are adjusted such that the number of conformations sampled per sequence and temperature is roughly the same (see text). }
\end{figure}

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two random sequences, R1 and R2, chosen at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, an MS simulation provides the thermodynamics of all 144 sequences at one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). We first validate the MS algorithm by comparing the average total energy, $\left < E \right >$, calculated for these 4 sequences with the two different methods (see Supplementary Information). The two sets of results are entirely consistent showing that, for a given $\sbar$ and $T$, both MS and ST indeed sample the same distribution.

As a way to assess conformational sampling efficiency, we compare in Fig.~4 the statistical error, $\sigE$, of the average energy $\left <E\right >$ for the 4 sequences obtained using ST and MS, respectively. Because approximately the same number of sampled conformations were obtained for each combination of $\sbar$ and $T$, we compare the statistical errors directly. At the highest studied $T$, which is well above the folding temperature of both A1 and N1, the two algorithms give almost identical statistical errors. This can be understood by noting that at high-$T$ the free-energy landscape is relatively smooth and conformational space requires little difficulty to sample. The benefit of adding a dynamic parameter, whether $\sbar$ or $T$, is apparently minor under these conditions. However, at lower $T$, the $\sigE$ values from MS is often smaller than those from ST and never significantly higher. For example, at the lowest $T$, the precision in the estimate of $\left < E\right >$ is roughly twice as high in MS than ST for A1 and R1, and roughly the same for N1 and R2.

@@

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{network}}
\caption{Networks of sequences connecting folds IA and IB (top) and folds IIA and IIB (bottom). Each node represents a stable sequence ($\Ptot\ge\Pcut$ where $\Pcut=0.50$) that folds into either IA or IIA (light blue), IB or IIB (dark red), or is classified as bistable ($B>0.5$, black). A line between two nodes indicates that the sequences differ at only one position. Graph created using the tool Graphviz~\protect\cite{Graphviz2000} obtained from  www.graphviz.org.}
\end{figure*}

@@

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to abrupt IA-IB or IIA-IIB fold changes, i.e., paths that do not pass through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at a single amino acid position. To determine if a sequence is stable we use the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PIA + \PIB$ and $\PIIA + \PIIB$ for the IA-IB and IIA-IIB fold pairs, respectively; $\Ptot$ thus indicates the total stability of a sequence across the two competing folds. The precise network depends, of course, on the cut-off value of $\Pcut$ and a higher $\Pcut$ generally means a selection of more stable pathways.

>>>>

\begin{abstract}
We consider a generalized-ensemble algorithm for coarse-grained simulations of biomolecules which allows the thermodynamic behavior of two or more sequences to be determined in a single multisequence run. By carrying out a random walk in sequence space, the method also enhances conformational sampling. Escape from local energy minima is accelerated by visiting sequences for which the minima are more shallow or absent. We test the method on an intermediate-resolution coarse-grained model for protein folding with 3 amino acid types and explore the potential for large-scale coverage of sequence space by applying the method to sets of more than 1,000 sequences. The resulting thermodynamic data is used to analyze the structures and stability properties of sequences covering the space between folds with different secondary structures. 
\end{abstract}

@@

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, many biologically motivated problems naturally call for the investigation and comparison of molecular variants, e.g., determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} and protein-nucleotide~\cite{Rohs2010} interactions, or the role of mutations in molecular disease processes.~\cite{Ross2004} Another example is protein folding, where unique insight has been achieved by comparing sequences within and between protein families.~\cite{Tzul2017,Wensley2010} In a situation with extremely rapid growth of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

To this end, we consider in this work an MC-based algorithm that can calculate the thermodynamics of multiple sequences in a single run and apply it to a coarse-grained model for protein folding.~\cite{Bhattacherjee2012} This multisequence (MS) method was originally developed in the context of homo- and heteropolymer simulations~\cite{Irback1995} and was later adapted for the characterization of peptide-protein binding specificity.~\cite{Bhattacherjee2013,Wallin2017} To our knowledge, it has not been previously tested in realistic protein folding simulations. The MS algorithm carries out a simulation in a generalized ensemble that performs a random walk in sequence space. Hence, there are two main types of updates: conformational updates $\rbar\rightarrow\rbar'$ and sequence updates $\sbar\rightarrow\sbar'$. This strategy is straightforward when $\rbar$ and $\sbar$ are ``perpendicular" coordinates, as illustrated in Fig.~1,  such that the potential energy of the model can be written in terms of two independent variables, $E(\sbar,\rbar)$.

%While this may be hard to achieve for detailed biomolecular models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model protein sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and a $\beta$-hairpin,~\cite{Bhattacherjee2012} respectively, as shown in Fig.~2. In a previous work, this set allowed us to demonstrate the existence of A1-N1 mutational pathways which do not proceed through any disordered (non-folding) sequences but instead switch abruptly between the two folds in a single mutational step.~\cite{Holzgrafe2014} Such an ability to reversibly switch folds was recently demonstrated in a handful of natural and engineered proteins.~\cite{Bryan2010} Here we use the set {\SR} and our CG model for protein folding to validate the MS method and compare its efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences, also spanning the A1-N1 space, as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively (see Fig.~2). Besides demonstrating that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

@@

\subsection{Generalized-ensemble algorithms and simulated tempering}
\noindent
Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters.

A well-known generalized-ensemble algorithm is simulated tempering (ST),~\cite{Marinari1992,Lyubartsev1992} where it is the temperature that becomes a dynamic parameter. In this scheme, frequent visits to high $T$ allow simulations to readily escape from local energy minima. The ST algorithm simulates the joint probability distribution 
\begin{equation}
P(\rbar,m) =\dfrac{1}{\hat{Z}} e^{-\beta_m E(\rbar) + g_m}\,,
\label{ST}
\end{equation}
where  $\beta_m=1/k_\mathrm{B} T_m$, $\{T_m\}_{m=1}^\mathrm{M}$ a set of temperatures and $k_\mathrm{B}$ is Boltzmann's constant. The normalization constant in Eq.~\ref{ST} is  
\begin{equation}
\hat{Z} = \sum_r \sum_{m=1}^{\mathrm{M}}e^{-\beta_m E(\rbar) + g_m}\,,
\end{equation}
where the first sum is over all conformations $\rbar$. The simulation parameters $g_m$ control the marginal probability distribution
\begin{equation}
P(m) = \frac{1}{\hat{Z}}\sum_r e^{-\beta_m E(\rbar) + g_m} \,,
\end{equation}
and must therefore be carefully chosen. A common and convenient choice is $g_m\approx \beta_m F_m$, where $F_m$ is the free energy at temperature $T_m$. With this choice, $P(m)$ becomes approximately flat ensuring all temperatures are frequently visited.

\subsection{Multisequence algorithm}
\noindent 
The basic idea of the MS algorithm for biomolecular simulation is to let the sequence $\sbar$ become a dynamic parameter rather than the temperature as in ST. A dynamic $\sbar$ is technically feasible when the potential energy can be written as $E(\sbar,\rbar)$, where $\sbar$ and $\rbar$ are independent variables. This is the case in our coarse-grained protein model which has only backbone degrees of freedom. It can also be achieved for in some more detailed models.~\cite{Bhattacherjee2013,Wallin2017}

Similarly to ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{Z}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
Z = \sum_{\sbar}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\end{equation}
and the first sum goes over a set of allowed sequences $\sbar$. The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=Z^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

@@

This geometrical description is paired with a simplified but finely tuned energy function with 4 terms: $E= E_\mathrm{ev} + E_\mathrm{loc} + E_\mathrm{hb} + E_\mathrm{hp}$. The first two, $E_\mathrm{ev}$ and $E_\mathrm{loc}$, represent excluded-volume effects and local electrostatic effects, respectively. The hydrogen-bond energy, $E_\mathrm{hb}$, represents directionally dependent interactions between NH and CO groups and is necessary for secondary structure formation. Finally, the ``hydrophobicity" term, $E_\mathrm{hp}$, implements pairwise Lennard-Jones-like interactions between the $C_\beta$ atoms of h amino acids which are necessary for driving chain collapse during folding. Various model parameters, e.g., the strengths of hydrophobic attractions and hydrogen bonding, were determined based on the ability of the model to spontaneously fold a set of model sequences with 18-54 amino acids into  structurally diverse and thermodynamically stable native states with both $\beta$ and $\alpha$-structure. As it turned out, this strategy made the model robust enough to fold sequences designed to have mixed $\alpha$ and $\beta$ structures.

@@

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and described briefly below. First we define two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating the fraction of the fold-specific backbone-backbone hydrogen bonds that have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The stabilities of folds IA and IB are defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.8)$, respectively, i.e., the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ thus depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.875\pm 0.003$ for A1 and $\PIB=0.785\pm0.008$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $\mathrm{C}_\beta$-$\mathrm{C}_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} In analogy with $\PIA$ and $\PIB$, we define the stabilities of folds IIA and IIB as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.8)$, respectively. The root-mean-square-deviation, RMSD, is calculated over all $\mathrm{C}_\alpha$ atoms.

\begin{figure}
\includegraphics[width=8.3cm]{Fig2}
\caption{(A) Example of an MS simulation applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$. (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

@@

\subsection{Computational efficiency}
\noindent
We start by applying the MS algorithm to the set {\SR} across a range of temperatures $T$ (see Table~II). Two of the sequences in {\SR} are A1 and N1 (see Table~I) which fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. Because A1 and N1 differ at 10 positions, 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences in {\SR} were selected from this binary space with the constraints that the total number of hydrophobic amino acids are not too high and that they are not too unevenly distributed along the sequence.~\cite{Holzgrafe2014}

\begin{figure}
\includegraphics[width=7.8cm]{Fig3}
\caption{Acceptance rates for $\sbar\rightarrow\sbar'$ updates in MS simulations of the {\SR} sequence set as a function of (A) the number of changed amino acid positions $\Delta h$ and (B) temperature $T$. Acceptance rates for 3 different $T$'s are shown in (A).}
\end{figure}

Figure~2 illustrates a typical MS simulation trajectory carried out at the lowest studied temperature which is below the folding temperature of both A1 and N1.~\cite{Holzgrafe2014,Holzgrafe2015} From the MC evolution of the total energy $E$, sequence index $\sbar$, and RMSD values from the representative structures in Fig.~2B, it is evident that visits to various sequences drive transitions into a range of structural states. In particular, there are frequent visits to $\alpha$-helix and $\beta$-hairpin structures and transitions between them are accompanied by a shift in which sequences are preferably visited. For example, visits to high $\sbar$-indices, including N1 with index 144, tend to coincide with formation of $\beta$-hairpin structures as required to generate correct equilibrium conformational ensembles.

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rate is substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~4B). An increased acceptance rate can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains.

\begin{figure}
\includegraphics[width=8.0cm]{Fig4}
\caption{Comparing sampling efficiency of the MS and ST algorithms. Statistical errors $\sigma_{\left < E\right >}$ of the average total energy $\left < E\right >$ from the two methods for the 16-amino acid sequences (A) A1, (B) N1, (C) R1 and (D) R2. Simulation lengths in the two methods are adjusted such that the number of conformations sampled per sequence and temperature is roughly the same (see text). }
\end{figure}

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two random sequences, R1 and R2, chosen at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, an MS simulation provides the thermodynamics of all 144 sequences at one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). We first validate the MS algorithm by comparing the average total energy, $\left < E \right >$, calculated for these 4 sequences with the two different methods (see Supplementary Information). The two sets of results are entirely consistent showing that, for a given $\sbar$ and $T$, the MS and ST algorithms indeed sample the same distribution.

As a way to assess conformational sampling efficiency, we compare in Fig.~4 the statistical error, $\sigE$, of the average energy $\left <E\right >$ for the 4 sequences obtained using ST and MS, respectively. Because approximately the same number of sampled conformations were obtained for each combination of $\sbar$ and $T$, we compare the statistical errors directly. At the highest studied $T$, which is well above the folding temperature of both A1 and N1, the two algorithms give almost identical statistical errors. This can be understood by noting that at high-$T$ the free-energy landscape is relatively smooth and conformational space requires little difficulty to sample. The benefit of adding a dynamic parameter, whether $\sbar$ or $T$, is apparently minimal under these conditions. However, at lower $T$, the $\sigE$ values from MS is often smaller than those from ST and never significantly higher. For example, at the lowest $T$, the precision in the estimate of $\left < E\right >$ is roughly twice as high in MS than ST for A1 and R1, and roughly the same for N1 and R2.

@@

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{Fig5}}
\caption{Networks of sequences connecting folds IA and IB (top) and folds IIA and IIB (bottom). Each node represents a stable sequence ($\Ptot\ge\Pcut$ where $\Pcut=0.50$) that folds into either IA or IIA (light blue), IB or IIB (dark red), or is classified as bistable ($B>0.5$, black). A line between two nodes indicates that the sequences differ at only one position. Graph created using the tool Graphviz~\protect\cite{Graphviz2000} obtained from  www.graphviz.org.}
\end{figure*}

@@

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to abrupt IA-IB or IIA-IIB fold changes, i.e., paths that do not pass through any unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at a single amino acid position. To determine if a sequence is stable we use the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PIA + \PIB$ and $\PIIA + \PIIB$ for the IA-IB and IIA-IIB fold pairs, respectively; $\Ptot$ thus indicates the total stability of a sequence across the two competing folds. The precise network depends, of course, on the cut-off value $\Pcut$ and a higher $\Pcut$ generally means a selection of more stable pathways.

Fig.~5 illustrates the networks obtained with $\Pcut=0.50$ showing that both the IA-IB and IIA-IIB fold pairs are connected in sequence space at this stability threshold. A precise analysis shows that there are 516,972 viable IA-IB paths and 57,912 viable IIA-IIB paths. These paths represent 14.2 \% and 1.6 \% of all possible paths, respectively, because there are a total of $10! =3,628,800$ possible paths between start and end points in both cases. Hence, folds IA and IB are rather highly connected in our model for $\Pcut= 0.50$. For $\Pcut=0.60$, the numbers are 104,640 paths (2.9\%) for IA-IB and 22,512 (0.6\%) paths for IIA-IIB. We find that there are no possible IA-to-IB or IIA-to-IIB paths when $\Pcut\ge0.74$ and $\ge 0.66$, respectively. \\
